---
title: "Rbootcamp_final"
author: "Yvonne Boadi"
format: 
  html:
   theme: yeti
execute: 
  echo: false 
  error: true
---
```{r}
library(tidyverse)
library(gtsummary)
library(dplyr)
library(readr)
health_spending <- read_csv("~/RSPH Classes/EPI590R RStudio/final-project/final-project/repository/health_spending.csv")
View(health_spending)
```
```{r}
#The data tracks how much countries spend on health care, where the money comes from (government, private, or external sources), how it is channelled through financing schemes (e.g. government programmes, voluntary insurance, out-of-pocket payments), and what it is spent on (e.g. curative care, preventive care, medical goods). How health systems are funded and how resources are allocated directly shapes health outcomes, progress toward universal health coverage, and whether households face financial hardship when accessing care. These indicators are essential to inform health policy decisions.
```

```{r}
#| label: tbl-one
#| tbl-cap: "My Table Caption"
#| message: true
#| warning: true


health_spending_filtered <- health_spending %>%
  filter(country_name %in% c("Guyana", "Haiti")) 

health_spending_wide <- health_spending_filtered |>
	pivot_wider( #takes values from one column and spreads them out into new columns, using another column to supply the new column names
		id_cols = c(country_name, year), #
		names_from = indicator_code,
		values_from = value
	) |>
	mutate ( year_centered = year - (mean(year)))


table1 <- tbl_summary(
  health_spending_wide, 
  by = country_name, 
  include = c(year, ext_che, gghed_che, pvtd_che),
  label = list(
    ext_che ~ "External aid %",
    gghed_che ~ "Government aid %",
    pvtd_che ~ "Private aid %"
  ),
  missing_text = "Missing"
)
table1


#?inline_text()

```

@tbl-one shows the descriptive statistics about Guyana's and Haiti's aid use (shows the median and IQR)

This table shows the median and interquile ranges for the external, government and private aid that Guyana and Haiti utlizes from 2000 to 2023. The values on this table represents the median and IQR. The median and IQR for Guyana's external aid is `r inline_text(table1, variable = ext_che, column = "Guyana")`


```{r}
#| label: fig-scatterplot
#| fig-cap: "This scatterplot shows the relationship between Guyana and Haiti's external aid use (in percentage)"

library(ggplot2) 

#Create the scatter plot 



ggplot (data = health_spending_wide, mapping = aes (x = year, y = ext_che, color = country_name)) + geom_point(size = 3) +
	labs (title = "External Aid Share of Health Spending Over Time",
				x = "Years",
				y = "External Share of CHE (%)",
				color = "Country Name") +
	theme_minimal()


```
@fig-scatterplot shows the years and external share of CHE in percentage and @fig-regression indicates that overtime, Haiti's external aid utilization increased.

```{r}
#| label: fig-regression
#| fig-cap: "The regression analysis shows the relationship between external aid use (in percentage) between Haiti and Guyana."

linear_model <- (lm(ext_che ~ year_centered * country_name, data = health_spending_wide))

tbl_regression(linear_model, intercept = TRUE,
							 label = list(year_centered ~ "Year (trend)",
							 						 country_name ~ "Country",
							 						 `year_centered:country_name` ~ "Trend difference (Haiti vs Guyana"))

ggplot(health_spending_wide, aes(x = year_centered, y = ext_che, color = country_name)) +
	geom_point(alpha = 0.4) + #draws the actual data as dots, positioning the x & y. alpha controls the transparency
	geom_smooth(method = "lm", se = TRUE) + #visusallizes your regression. method = fits the straight line linear regression. se = is standard error
	labs(
		title = "External Aid Share of Health Spending Over Time",
		x = "Year", y = "External Share of CHE (5)",
		color = "Country"
	)
```


```{r}

lin_reg <- tbl_regression(linear_model, intercept = TRUE,
							 label = list(year_centered ~ "Year (trend)",
							 						 country_name ~ "Country",
							 						 `year_centered:country_name` ~ "Trend difference (Haiti vs Guyana"))
lin_reg

coef_marker <- inline_text(lin_reg, variable = "(Intercept)", pattern = "{estimate} ({conf.low}, {conf.high})")
coef_marker

```


